---
title: "European Gas Trade Flows 2022"
author: "Jorge Bandeira"
date: "`r Sys.Date()`"
output: html_document
---

## 1. Summary
The following analysis aims to display the main trade flows of natural gas in Europe.
The data used for this analysis was downloaded from the [IEA website](https://www.iea.org/data-and-statistics/data-product/gas-trade-flows).

## 2. Data processing
First, the dataset was imported into an R dataframe and converted to a long format unpivoting the month columns.

``` {r eval = FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(data.table)
library(lubridate)

# read raw xls file
raw_df <- read_excel("Export_GTF_IEA_202205_edited.xls")

#unpivot months
unp_df <- melt(setDT(raw_df), id.vars = c("Borderpoint", "Exit", "Entry"), variable.name = "Month")

```

Next, the original "Month" values were converted to dates and separate columns for month and year were created.

``` {r eval = FALSE, results='hide'}
#create year and month columns
unp_df <- unp_df %>%
  transform(Month = as.numeric(as.character(Month))) %>% 
  mutate(date = as_date(Month, origin = date("1900/01/01"))) %>% 
  mutate(month = month(date)) %>% 
  mutate(year = year(date)) %>% 
  select(-c(Month, date))
```

It was detected that some flow values were displayed as "N/A". These values were switched to 0 and the data type of the flow column was changed to numeric to allow calculations to be performed.
Column names we also changed to lower case for consistency. The clean data was then exported to a csv file.

``` {r eval = FALSE, results='hide'}
#change NA values to 0, make gtf column numeric
unp_df$gtf_Mcm[unp_df$gtf_Mcm == "N/A"] <- 0
unp_df <- unp_df %>% 
  mutate(gtf_Mcm = as.numeric(gtf_Mcm))

#change colnames to lowercase
names(unp_df) <- tolower(names(unp_df))

#export to csv
write_csv(unp_df, "gas_exports_data_cleaned.csv")
  
```

## 3. Visualizing the data

To setup the data for visualization, first, a new dataframe was created by filtering values from 2022, excluding LNG flows since those are out of scope and grouping the data by exit and entry pairs.

``` {r eval = TRUE, results='hide'}
library(tidyverse)
library(networkD3)

#import clean data
gas_data <- read_csv("gas_exports_data_cleaned.csv")

#function to get top exporters for a given year
getTrade <- function(f_year) {
    gas_data %>%
    select(exit, entry, gtf_mcm, year) %>% 
    filter(year == f_year, entry != "Liquefied Natural Gas", exit != "Liquefied Natural Gas") %>% 
    group_by(exit, entry) %>% 
    summarise(export = sum(gtf_mcm)) %>% 
    arrange(desc(export)) %>% 
    filter(export > 2000)
}

top_exp_2022 <- getTrade(2022)
```

To visualize the data, it was decided that a good option would be to use a sankey network style chart.
The library "networkD3" was used.
To allow the creation of the chart "networkD3" requires a dataframe with the unique nodes as well as unique Ids for each node.

``` {r eval = TRUE, results='hide'}
#create nodes df for networkD3
nodes_2022 <- data.frame(name=c(as.character(top_exp_2022$exit), as.character(top_exp_2022$entry)) %>% unique())

#set ids for networkD3
top_exp_2022$exitId = match(top_exp_2022$exit, nodes_2022$name) - 1
top_exp_2022$entryId = match(top_exp_2022$entry, nodes_2022$name) - 1

#create chat with overview of natural gas trade
gas_trade_chart_2022 <- sankeyNetwork(Links = top_exp_2022, Nodes = nodes_2022, Source = "exitId", Target = "entryId", 
                                      Value = "export", NodeID = "name", sinksRight = TRUE, fontFamily = "Arial", fontSize = 10, nodePadding = 20)


```
```{r message=FALSE, results='hold', fig.align='center'}
sankeyNetwork(Links = top_exp_2022, Nodes = nodes_2022, Source = "exitId", Target = "entryId", 
                                      Value = "export", NodeID = "name", sinksRight = TRUE, fontFamily = "Arial", fontSize = 10, nodePadding = 20)
```

The visual displays the main natural gas trade flows (> 2000 Mcm) in Europe.
